Compact orbitals enable low-cost linear-scaling \emph{ab initio} molecular dynamics for weakly-interacting systems

### 2018-05-26 JCP/manuscript.tex
JCP: update

\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{amsmath}% bold math
\usepackage{siunitx}% si units
\usepackage{color}
\usepackage{graphicx}
%\usepackage[normalem]{ulem}
%\nofiles

@@

@@

ALMO AIMD was implemented in CP2K, an open source materials modeling package~\cite{www:cp2k}. 
Accuracy and efficiency of ALMO AIMD was tested using liquid water as an example. 
%RZK: cite our own water paper
This system is challenging because intermolecular electron delocalization is a critical component of hydrogen bonding and must be described correctly to reproduce static and dynamical properties of liquid water. 
%To calculate properties of water at ambient conditions, 
A periodic cell containing 125 molecules was simulated for $\SI{30}{\ps}$ at $T=\SI{298}{\K}$ and a constant density of $\SI{1.01}{\g\cdot\cm^{-3}}$. 
The initial configurations of the water boxes were created by equilibrating the system for 20~ps using the fully converged orbitals. 
Ricci-Ciccotti algorithm~\cite{Ricci2003} was used to integrate the Langevin equation (see Supporting Information). We found that $\gamma = \SI{e-3}{\per\fs}$ is large enough to thermostat the system efficiently and small enough not to significantly affect dynamical properties of liquid water.
%
In the dual Gaussian and plane-wave scheme implemented in CP2K~\cite{a:quickstep}, the TZV2P basis set was used to represent molecular orbitals, and a plane-wave cutoff of $\SI{320}{Ry}$ used to represent electron density. 
The XC energy was approximated using the dispersion-corrected PBE functional~\cite{a:pbe,Grimme2010}. 
Separable norm-conserving pseudopotentials were used~\cite{a:hgh} and the Brillouin zone was sampled at the $\Gamma$-point. 
%Although not a required component of ALMO AIMD, 
The predictor of the Kolafa scheme~\cite{Kolafa2003} was adopted to localized orbitals~\cite{a:2ndcpmd} to generate a highly accurate initial ALMOs in both SCF stages, which can be brought close to the ground state with just a few SCF steps of the robust two-stage optimization procedure. 
%RZK: The time-reversibility of ASPC is not a critical reqirement of the proposed method.

@@

Even with $\epsilon_{\text{SCF}} = 10^{-2}$, the simulation is stable with the correct average temperature and perfect Maxwell-Boltzmann distribution (Figure~\ref{fig:dynproperties}b). $\Delta$ was initially estimated at $\SI{2e-5}{\per\fs}$ using Eq.~(\ref{eq:delta}) and then refined heuristically to $\SI{6e-5}{\per\fs}$. 
We found that it is easier to optimize $\Delta$ when $\gamma$ is set to zero because of reduced noise in the trial runs. 
%RZK addition for a longer article: In theory, $\Delta$ should be independent on system size. However, we found that $\Delta$ dependents on system size due to finite size effects, which decay for increasing system size.
%
Analysis of $\delta \vec{f_{i}}(t)$ shows that the error indeed resembles Gaussian white noise. The mean of the error is zero (black line in Figure~\ref{fig:randomforce}a). Its ACF decays rapidly (Figure~\ref{fig:randomforce}c) so that the errors can be considered uncorrelated on time scale of $\SI{50}{\fs}$. Thus the main assumption behind our approach to ALMO AIMD is justified for liquid water. We established that the main source of error in forces for this system is the loose convergence criterion and not the finite $R_c$: fully converged ALMO SCF calculations remove the oscillating component of $\delta f$ (Figure~\ref{fig:randomforce}b). We also verified that the ALMO forces converge to the reference forces in the limit $R_{c} \rightarrow \infty$ 
%(Figure~\ref{fig:forcecomp} in the Supplemental Material).
(Figure~SZZZ in the Supporting Information).

@@

% RZK longer article: averaging is done over 10 time steps 
To demonstrate the computational efficiency of ALMO AIMD, we compared the average wall-time per MD step for a variety of methods in Figure \ref{fig:strongscaling_log}.
%In addition to the fully converged OT reference, we included timing for the second generation Car-Parrinello method designed to decrease the computational prefactor of the cubically-scaling DFT~\cite{a:2ndcpmd}. 
It is important to emphasize that the comparison is performed for a three-dimensional condensed phase system described with an accurate triple-$\zeta$ basis set with polarized functions---a particularly challenging case for DM-based LS methods.
ALMO AIMD shows clear LS behavior for all values of $\epsilon_{\text{SCF}}$, even for medium-size systems. While the second generation Car-Parrinello method decreases the computational overhead of the cubically-scaling AIMD for small systems~\cite{a:2ndcpmd}, ALMO AIMD exploits the modified Langevin concept to substantially reduce the simulation cost for systems of all sizes.
The crossover point between ALMO AIMD and cubically scaling methods lies in the region of 256 molecules---length scales routinely accessible with AIMD today. For comparison, the crossover point between the cubically-scaling DFT and a representative LS method that computes a sparse DM as the matrix sign function of the effective Hamiltonian~\cite{a:ls-dm-sign} is expected to lie well above 10,000 water molecules (Figures 5a and S5 in Ref.~\onlinecite{:a:almo-ls}).

>>>>

\usepackage{xr}
\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{amsmath}% bold math
\usepackage{siunitx}% si units
\usepackage{color}
\usepackage{graphicx}
%\usepackage[normalem]{ulem}
%\nofiles

@@

\externaldocument[SI-]{SI}

@@

ALMO AIMD was implemented in CP2K, an open source materials modeling package~\cite{www:cp2k}. 
Accuracy and efficiency of ALMO AIMD was tested using liquid water as an example. 
This system is challenging because intermolecular electron delocalization is a critical component of hydrogen bonding and must be described correctly to reproduce static and dynamical properties of liquid water. 
A periodic cell containing 125 molecules was simulated for $\SI{30}{\ps}$ at $T=\SI{298}{\K}$ and a constant density of $\SI{1.01}{\g\cdot\cm^{-3}}$. 
The initial configurations of the water boxes were created by equilibrating the system for 20~ps using the fully converged orbitals. 
Ricci-Ciccotti algorithm~\cite{Ricci2003} was used to integrate the Langevin equation (see Supporting Information). We found that $\gamma = \SI{e-3}{\per\fs}$ is large enough to thermostat the system efficiently and small enough not to significantly affect dynamical properties of liquid water.
%
In the dual Gaussian and plane-wave scheme implemented in CP2K~\cite{a:quickstep}, the TZV2P basis set was used to represent molecular orbitals, and a plane-wave cutoff of $\SI{320}{Ry}$ used to represent electron density. 
The XC energy was approximated using the dispersion-corrected PBE functional~\cite{a:pbe,Grimme2010}. 
Separable norm-conserving pseudopotentials were used~\cite{a:hgh} and the Brillouin zone was sampled at the $\Gamma$-point. 
%Although not a required component of ALMO AIMD, 
The predictor of the Kolafa scheme~\cite{Kolafa2003} was adopted to localized orbitals~\cite{a:2ndcpmd} to generate a highly accurate initial ALMOs in both SCF stages, which can be brought close to the ground state with just a few SCF steps of the robust two-stage optimization procedure. 
%RZK: The time-reversibility of ASPC is not a critical reqirement of the proposed method.

@@

Even with $\epsilon_{\text{SCF}} = 10^{-2}$, the simulation is stable with the correct average temperature and perfect Maxwell-Boltzmann distribution (Figure~\ref{fig:dynproperties}b). $\Delta$ was initially estimated at $\SI{2e-5}{\per\fs}$ using Eq.~(\ref{eq:delta}) and then refined heuristically to $\SI{6e-5}{\per\fs}$. 
We found that it is easier to optimize $\Delta$ when $\gamma$ is set to zero because of reduced noise in the trial runs. 
%RZK addition for a longer article: In theory, $\Delta$ should be independent on system size. However, we found that $\Delta$ dependents on system size due to finite size effects, which decay for increasing system size.
%
Analysis of $\delta \vec{f_{i}}(t)$ shows that the error indeed resembles Gaussian white noise. The mean of the error is zero (black line in Figure~\ref{fig:randomforce}a). Its ACF decays rapidly (Figure~\ref{fig:randomforce}c) so that the errors can be considered uncorrelated on time scale of $\SI{50}{\fs}$. Thus the main assumption behind our approach to ALMO AIMD is justified for liquid water. We established that the main source of error in forces for this system is the loose convergence criterion and not the finite $R_c$: fully converged ALMO SCF calculations remove the oscillating component of $\delta f$ (Figure~\ref{fig:randomforce}b). We also verified that the ALMO forces converge to the reference forces in the limit $R_{c} \rightarrow \infty$ 
%(Figure~\ref{fig:forcecomp} in the Supplemental Material).
(Figure~\ref{SI-fig:forcecomp} in the Supporting Information).

@@

% RZK longer article: averaging is done over 10 time steps 
To demonstrate the computational efficiency of ALMO AIMD, we compared the average wall-time per MD step for a variety of methods in Figure \ref{fig:strongscaling_log}.
%In addition to the fully converged OT reference, we included timing for the second generation Car-Parrinello method designed to decrease the computational prefactor of the cubically-scaling DFT~\cite{a:2ndcpmd}. 
It is important to emphasize that the comparison is performed for a three-dimensional condensed phase system described with an accurate triple-$\zeta$ basis set with polarized functions---a particularly challenging case for DM-based LS methods.
ALMO AIMD shows clear LS behavior for all values of $\epsilon_{\text{SCF}}$, even for medium-size systems. While the second generation Car-Parrinello method decreases the computational overhead of the cubically-scaling AIMD for small systems~\cite{a:2ndcpmd}, ALMO AIMD exploits the modified Langevin concept to substantially reduce the simulation cost for systems of all sizes.
The crossover point between ALMO AIMD and cubically scaling methods lies in the region of 256 molecules---length scales routinely accessible with AIMD today. For comparison, the crossover point between the cubically-scaling DFT and a representative LS method that computes a sparse DM as the matrix sign function of the effective Hamiltonian~\cite{a:ls-dm-sign} is expected to lie well above 10,000 water molecules (Figures 5a and S5 in Ref.~\onlinecite{a:almo-ls}).
